\ttitle

### 2015-08-18 Chapters/CPET_Outcomes.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

\section{Introduction}
Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years. \parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999} In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy\parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.

The decision to operate on these patients depends not only on preoperative tumour stage but also on patient factors.\parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010} Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery. \parencite{mann_review_2010, mayo_management_2012} However, major pancreatic surgery is associated with significant morbidity and mortality and patients who have postoperative complications are less likely to get adjuvant therapy.\parencite{teh_patient_2009}

There have been a number of attempts to objectively define patient fitness and its relationship with postoperative outcome. Copeland and co-workers (1991) reported that the Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) criteria, in particular the POSSUM physiology score (PPS) could be used to quantify the risk of postoperative morbidity and mortality.\parencite{copeland_possum:_1991} However, the role of POSSUM in predicting postoperative outcome after surgery for pancreatic cancer is not entirely clear.\parencite{de_castro_evaluation_2009, khan_evaluation_2003, kocher_risk-adjustment_2005, pratt_possum_2008, tamijmarane_application_2008} The physiological component of POSSUM as well as other similar risk scoring systems such as E-PASS (Estimation of Physiologic Ability and Surgical Stress)\parencite{haga_estimation_1999} are calculated based on known comorbidities, clinically evident abnormalities in patient physiology or blood tests.

More recently, there has been some evidence that the presence of an ongoing systemic inflammatory response before surgery is associated with the development of postoperative complications in patients undergoing surgery for colorectal cancer\parencite{moyes_preoperative_2009}, oesophageal cancer\parencite{vashist_glasgow_2010} as well as pancreatic cancer.\parencite{knight_evaluation_2010}

Older and co-workers (1993) reported that cardiopulmonary exercise testing (CPET) was an objective evaluation of the response of the cardiovascular and respiratory systems to an increase in oxygen demand during exercise and was useful in predicting perioperative morbidity and mortality in patients undergoing major abdominal surgery.\parencite{older_preoperative_1993}

@@

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). An electrically-braked cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons. The VO$_2$AT was calculated using the V-slope\parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents\parencite{sue_metabolic_1988} methods. Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers\parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.

@@

Postoperative adverse events were recorded using internationally recognised definitions. The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae\parencite{bassi_postoperative_2005} and post-operative haemorrhage\parencite{wente_postpancreatectomy_2007}. The Clavien-Dindo (CD) classification\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major. Multiple admissions to critical care as well as re-operations were recorded. Operative mortality was defined as postoperative death in-hospital regardless of duration of stay or occurring within 30 days of the surgery. All complications were discussed at a weekly multidisciplinary meeting attended by three pancreatic surgeons and a radiologist with a specialist interest in pancreatic diseases and recorded in a prospective database.

@@

The incidence of pancreatic fistula was greater in patients with a low VO$_2$AT. This association needs further evaluation taking into consideration other well-recognised risk factors for pancreatic fistula such as pancreatic texture, pancreatic duct size and intra-operative blood loss.\parencite{braga_prognostic_2011, pratt_possum_2008, winter_1423_2006} It is possible that local or operative factors may be compounded by poor oxygen delivery and organ perfusion as measured by cardiopulmonary exercise testing. There was a non-significant trend towards clinically relevant pancreatic fistulae (ISGPS Grades B and C) as well as a significant association with major intra-abdominal abscesses (Clavien-Dindo Grades 3-5 i.e., requiring intervention, associated with organ dysfunction requiring intensive care or resulting in mortality). This would suggest that complications in patients with low VO$_2$AT are more likely to be severe than in patients with normal VO$_2$AT. However, there was no difference in mortality between patients with normal or low VO$_2$AT, indicating that multiple factors including preoperative patient fitness, local and operative factors, systemic inflammatory response, number of complications as well as perioperative critical care all play a role.

The results of this study also show that patients with a low VO$_2$AT were less likely to receive adjuvant therapy. Adjuvant therapy in patients undergoing pancreatic resections for cancer has been shown in multiple randomised trials to improve survival significantly.\parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009}
While postoperative mortality after pancreatic surgery has steadily improved over the years with major improvements in the quality of surgical and critical care over the past decade\parencite{winter_1423_2006} even in elderly patients\parencite{makary_pancreaticoduodenectomy_2006}, postoperative morbidity remains high.\parencite{mann_review_2010} The results of this study show that poor preoperative fitnees is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.

In the present study, VO$_2$AT was less than 10ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit. While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this.\parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010} Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993}

However, Snowden and co-workers\parencite{snowden_submaximal_2010} reported that patients with an VO$_2$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers\parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the VO$_2$AT did not predict postoperative adverse events less after oesophagectomy. The lack of association between low VO$_2$AT and cardiopulmonary complications in this study may have been due to two reasons. Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy.\parencite{braga_prognostic_2011} Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.

The results of this study are consistent with the results of the study by Ausania and co-workers\parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min. However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.

@@

This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’.\parencite{topp_effect_2002} It has been suggested that prehabilitation not only improves aerobic capacity\parencite{jones_effects_2007} but may also improve postoperative recovery.\parencite{mayo_impact_2011, pehlivan_effects_2011} The results of this study show that impaired aerobic capacity is associated with postoperative adverse events. Therefore, it would appear that prehabilitation using interventions such as exercise and nutrition, by improving physiological fitness, may have a role in improving postoperative outcomes after major pancreatic surgery and may improve the proportion of patients receiving adjuvant therapy.

>>>>

\section{Introduction}
Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years. \parencite{cancerresearchuk_cancer_2014}. The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999}. In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.

The decision to operate on these patients depends not only on preoperative tumour stage but also on patient factors \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}. Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery. \parencite{mann_review_2010, mayo_management_2012}. However, major pancreatic surgery is associated with significant morbidity and mortality and patients who have postoperative complications are less likely to get adjuvant therapy \parencite{teh_patient_2009}.

There have been a number of attempts to objectively define patient fitness and its relationship with postoperative outcome. Copeland and co-workers (1991) reported that the Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) criteria, in particular the POSSUM physiology score (PPS) could be used to quantify the risk of postoperative morbidity and mortality \parencite{copeland_possum:_1991}. However, the role of POSSUM in predicting postoperative outcome after surgery for pancreatic cancer is not entirely clear \parencite{de_castro_evaluation_2009, khan_evaluation_2003, kocher_risk-adjustment_2005, pratt_possum_2008, tamijmarane_application_2008}. The physiological component of POSSUM as well as other similar risk scoring systems such as E-PASS (Estimation of Physiologic Ability and Surgical Stress)\parencite{haga_estimation_1999} are calculated based on known comorbidities, clinically evident abnormalities in patient physiology or blood tests.

More recently, there has been some evidence that the presence of an ongoing systemic inflammatory response before surgery is associated with the development of postoperative complications in patients undergoing surgery for colorectal cancer \parencite{moyes_preoperative_2009}, oesophageal cancer \parencite{vashist_glasgow_2010} as well as pancreatic cancer \parencite{knight_evaluation_2010}.

Older and co-workers (1993) reported that cardiopulmonary exercise testing (CPET) was an objective evaluation of the response of the cardiovascular and respiratory systems to an increase in oxygen demand during exercise and was useful in predicting perioperative morbidity and mortality in patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.

@@

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). An electrically-braked cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons. The VO$_2$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods. Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.

@@

Postoperative adverse events were recorded using internationally recognised definitions. The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae \parencite{bassi_postoperative_2005} and post-operative haemorrhage \parencite{wente_postpancreatectomy_2007}. The Clavien-Dindo (CD) classification \parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major. Multiple admissions to critical care as well as re-operations were recorded. Operative mortality was defined as postoperative death in-hospital regardless of duration of stay or occurring within 30 days of the surgery. All complications were discussed at a weekly multidisciplinary meeting attended by three pancreatic surgeons and a radiologist with a specialist interest in pancreatic diseases and recorded in a prospective database.

@@

The incidence of pancreatic fistula was greater in patients with a low VO$_2$AT. This association needs further evaluation taking into consideration other well-recognised risk factors for pancreatic fistula such as pancreatic texture, pancreatic duct size and intra-operative blood loss \parencite{braga_prognostic_2011, pratt_possum_2008, winter_1423_2006}. It is possible that local or operative factors may be compounded by poor oxygen delivery and organ perfusion as measured by cardiopulmonary exercise testing. There was a non-significant trend towards clinically relevant pancreatic fistulae (ISGPS Grades B and C) as well as a significant association with major intra-abdominal abscesses (Clavien-Dindo Grades 3-5 i.e., requiring intervention, associated with organ dysfunction requiring intensive care or resulting in mortality). This would suggest that complications in patients with low VO$_2$AT are more likely to be severe than in patients with normal VO$_2$AT. However, there was no difference in mortality between patients with normal or low VO$_2$AT, indicating that multiple factors including preoperative patient fitness, local and operative factors, systemic inflammatory response, number of complications as well as perioperative critical care all play a role.

The results of this study also show that patients with a low VO$_2$AT were less likely to receive adjuvant therapy. Adjuvant therapy in patients undergoing pancreatic resections for cancer has been shown in multiple randomised trials to improve survival significantly \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009}
While postoperative mortality after pancreatic surgery has steadily improved over the years with major improvements in the quality of surgical and critical care over the past decade \parencite{winter_1423_2006} even in elderly patients \parencite{makary_pancreaticoduodenectomy_2006}, postoperative morbidity remains high \parencite{mann_review_2010}. The results of this study show that poor preoperative fitnees is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.

In the present study, VO$_2$AT was less than 10ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit. While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}. Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.

However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an VO$_2$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the VO$_2$AT did not predict postoperative adverse events less after oesophagectomy. The lack of association between low VO$_2$AT and cardiopulmonary complications in this study may have been due to two reasons. Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy \parencite{braga_prognostic_2011}. Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.

The results of this study are consistent with the results of the study by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min. However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.

@@

This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’ \parencite{topp_effect_2002}. It has been suggested that prehabilitation not only improves aerobic capacity \parencite{jones_effects_2007} but may also improve postoperative recovery \parencite{mayo_impact_2011, pehlivan_effects_2011}. The results of this study show that impaired aerobic capacity is associated with postoperative adverse events. Therefore, it would appear that prehabilitation using interventions such as exercise and nutrition, by improving physiological fitness, may have a role in improving postoperative outcomes after major pancreatic surgery and may improve the proportion of patients receiving adjuvant therapy.

### 2015-08-18 Chapters/CRP_Complications.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

Postoperative CRP levels have been shown to predict infective complications after major surgery including colorectal,  oesophago-gastric and pancreatic surgery. \parencite{mustard_c-reactive_1987, matthiessen_increase_2008, mcneer_early_2010, dutta_persistent_2011, mackay_c-reactive_2011, murakami_soft_2008, welsch_persisting_2008} It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.

@@

In the present study, while CRP was significantly greater in patients with POPF, this difference did not occur until the second postoperative day and therefore had no further clinical use in the prediction of this complication. The early rise in CRP in patients who developed a POPF is expected as the most common risk factors for POPF are encountered intra-operatively including soft pancreatic texture, small pancreatic duct size and operative blood loss. Soft pancreatic texture has in fact been reported to be associated with elevated CRP levels after pancreaticoduodenectomy.\parencite{murakami_soft_2008}

@@

The results of this study are similar to the findings of Welsch and co-workers who analysed 688 patients undergoing a variety of pancreatic resections with pancreaticojejunal anastomosis.\parencite{welsch_persisting_2008} They compared 91 patients who developed an 'inflammatory postoperative complication' with a random subset of 60 consecutive pancreatic resections. They reported that a threshold of 140 mg/L serum CRP level on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%). However, the definitions used for POPF in this study were different from the ISGPF definition and the authors did not report on the predictive ability of CRP in the presence or absence of a POPF. Moreover, Grade A and B POPF were not included in the analysis. Our results clearly demonstrate that early CRP elevation occurs in all grades of POPF with no difference between the different grades.

In the present study, preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score was not associated with infective complications. A preoperative acute phase response has been reported to be associated with an exaggerated postoperative systemic inflammatory response and septic complications. \parencite{haupt_association_1997} The modified Glasgow Prognostic Score, a validated measure of preoperative systemic inflammation, has also been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections.\parencite{moyes_preoperative_2009} The lack of association between the modified Glasgow Prognostic Score and complications in the present study may have been due to the multiple factors that are responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage procedures and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours. These findings are similar to those reported by Dutta and co-workers who did not find any relationship between preoperative systemic inflammation and postoperative complications in patients undergoing oesophago-gastric surgery.\parencite{dutta_persistent_2011}

The difference in white cell count and neutrophil count was not clinically significant between patients who did or did not have an infective complication. This finding is similar to one of the earliest reports on the role serial CRP measurements in predicting septic complications after surgery.\parencite{mustard_c-reactive_1987} as well as in several subsequent reports. \parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011} The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation. This study was not designed to look at inflammatory markers later in the postoperative period. However, Mustard and co-workers monitored CRP, white cell count and temperature for 14 days after surgery and found that CRP was the only reliable marker of septic complications.

@@

More recently, the compensatory anti-inflammatory response with the associated immunosuppressive effects has been recognised as a possible factor in predisposing the patient to infective complications and delayed wound healing. Such an anti-inflammatory response is proportional to the initial systemic inflammatory response and therefore patients with an exaggerated inflammatory response in the immediate aftermath of surgery may be at increased risk of developing complications. Mokart and co-workers studied trends in inflammatory cytokine levels in the serum of 30 patients undergoing major abdominal surgery for cancer. They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day. T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6.\parencite{faist_immunosuppression_1997} The complex interactions between the pro and anti-inflammatory processes, neurohormonal factors of the stress response, the role of the blood-gut barrier in bacterial translocation, local factors including tissue ischaemia and damage, the immunological effects of anaesthesia, blood transfusion and hypothermia remain the subject of ongoing research.\parencite{buttenschoen_effect_2010}

>>>>

Postoperative CRP levels have been shown to predict infective complications after major surgery including colorectal,  oesophago-gastric and pancreatic surgery \parencite{mustard_c-reactive_1987, matthiessen_increase_2008, mcneer_early_2010, dutta_persistent_2011, mackay_c-reactive_2011, murakami_soft_2008, welsch_persisting_2008}. It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.

@@

In the present study, while CRP was significantly greater in patients with POPF, this difference did not occur until the second postoperative day and therefore had no further clinical use in the prediction of this complication. The early rise in CRP in patients who developed a POPF is expected as the most common risk factors for POPF are encountered intra-operatively including soft pancreatic texture, small pancreatic duct size and operative blood loss. Soft pancreatic texture has in fact been reported to be associated with elevated CRP levels after pancreaticoduodenectomy \parencite{murakami_soft_2008}.

@@

The results of this study are similar to the findings of Welsch and co-workers who analysed 688 patients undergoing a variety of pancreatic resections with pancreaticojejunal anastomosis \parencite{welsch_persisting_2008}. They compared 91 patients who developed an 'inflammatory postoperative complication' with a random subset of 60 consecutive pancreatic resections. They reported that a threshold of 140 mg/L serum CRP level on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%). However, the definitions used for POPF in this study were different from the ISGPF definition and the authors did not report on the predictive ability of CRP in the presence or absence of a POPF. Moreover, Grade A and B POPF were not included in the analysis. Our results clearly demonstrate that early CRP elevation occurs in all grades of POPF with no difference between the different grades.

In the present study, preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score was not associated with infective complications. A preoperative acute phase response has been reported to be associated with an exaggerated postoperative systemic inflammatory response and septic complications \parencite{haupt_association_1997}. The modified Glasgow Prognostic Score, a validated measure of preoperative systemic inflammation, has also been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections \parencite{moyes_preoperative_2009}. The lack of association between the modified Glasgow Prognostic Score and complications in the present study may have been due to the multiple factors that are responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage procedures and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours. These findings are similar to those reported by Dutta and co-workers who did not find any relationship between preoperative systemic inflammation and postoperative complications in patients undergoing oesophago-gastric surgery \parencite{dutta_persistent_2011}.

The difference in white cell count and neutrophil count was not clinically significant between patients who did or did not have an infective complication. This finding is similar to one of the earliest reports on the role serial CRP measurements in predicting septic complications after surgery \parencite{mustard_c-reactive_1987} as well as in several subsequent reports \parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011}. The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation. This study was not designed to look at inflammatory markers later in the postoperative period. However, Mustard and co-workers monitored CRP, white cell count and temperature for 14 days after surgery and found that CRP was the only reliable marker of septic complications.

@@

More recently, the compensatory anti-inflammatory response with the associated immunosuppressive effects has been recognised as a possible factor in predisposing the patient to infective complications and delayed wound healing. Such an anti-inflammatory response is proportional to the initial systemic inflammatory response and therefore patients with an exaggerated inflammatory response in the immediate aftermath of surgery may be at increased risk of developing complications. Mokart and co-workers studied trends in inflammatory cytokine levels in the serum of 30 patients undergoing major abdominal surgery for cancer. They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day. T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6 \parencite{faist_immunosuppression_1997}. The complex interactions between the pro and anti-inflammatory processes, neurohormonal factors of the stress response, the role of the blood-gut barrier in bacterial translocation, local factors including tissue ischaemia and damage, the immunological effects of anaesthesia, blood transfusion and hypothermia remain the subject of ongoing research \parencite{buttenschoen_effect_2010}.
